---
title: "Plotting bipartite networks with ggbipart"
author: "Pedro Jordano"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting bipartite networks with ggbipart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

`ggbipart` is a set of `R` functions for plotting bipartite networks, using
`ggplot2`, `network`, and `igraph` graphics. The `ggplot2`-based routines rely
heavily on code developed by Francois Briatte for the
[`ggnet`](https://briatte.github.io/ggnet/) library (now part of `GGally`).

Bipartite networks are a special type of network where nodes belong to two
distinct sets (modes), and links only connect nodes of *different* sets. As
with other networks, bipartite structures can be **binary** (only the
presence/absence of a link is recorded) or **quantitative** (weighted), where
links carry a variable importance or weight. Everything starts from an
adjacency (incidence) matrix. Here we use matrices describing ecological
interactions between animal frugivores and the plants whose fruits they eat.
By convention in these examples, rows are plant species and columns are
animal species.

```{r packages, message = FALSE, warning = FALSE}
library(ggbipart)
library(ggplot2)
library(network)
library(igraph)
```

## The example data

The package ships two well-sampled plant–frugivore interaction matrices from
southern Spain in `inst/extdata/`, retrievable with `system.file()`. Both are
weighted adjacency matrices; the comment header lines are stripped with
`comment.char = "#"`.

```{r read_data}
# Nava de las Correhuelas (26 plants x 36 frugivores).
nch <- as.matrix(read.csv(
  system.file("extdata", "sdw01_adj_fru.csv", package = "ggbipart"),
  comment.char = "#", row.names = 1, check.names = FALSE))

# Hato Raton (16 plants x 17 frugivores).
hr <- as.matrix(read.csv(
  system.file("extdata", "sdw02_adj_fru.csv", package = "ggbipart"),
  comment.char = "#", row.names = 1, check.names = FALSE))

dim(nch)
dim(hr)
```

## Initializing bipartite networks

Adjacency matrices are turned into the graph objects the plotting functions
expect. `bip_init_network()` returns a `network` object; `bip_init_igraph()`
returns an `igraph` object. Both encode the two modes and carry the edge
weights.

```{r init_graphs}
hr.net <- bip_init_network(hr)   # network object
hr.ig  <- bip_init_igraph(hr)    # igraph object

hr.net
```

Two small helpers underlie the plotting routines. `vectorize()` converts an
adjacency matrix into a long, three-column edge table, and `bip_edgewt()`
returns log-scaled edge weights suitable for mapping to line widths.

```{r helpers}
head(vectorize(hr))
summary(bip_edgewt(hr, x = 30))
```

## Plotting with the `network` back end

`bip_qtplot()` draws a weighted (quantitative) bipartite network directly from
the adjacency matrix, scaling edge widths by interaction strength.

```{r qtplot}
bip_qtplot(hr)
```

`bip_binplot()` plots a `network` object; it is well suited to binary
(presence/absence) webs.

```{r binplot}
bip_binplot(hr, hr.net)
```

## Plotting with the `igraph` back end

`bip_igplot()` renders the same network through `igraph`, using the adjacency
matrix (for node counts and edge scaling) and the `igraph` object.

```{r igplot}
bip_igplot(hr, hr.ig)
```

## `ggplot2` graphs with `bip_ggnet`

`bip_ggnet()` builds the plot within the `ggplot2` framework via
`GGally::ggnet2`. `ggnet2` automatically detects two-mode graphs and
understands arguments of the form `[color, shape, size] = "mode"`, mapping the
two modes to the `ggnet2` classes. The returned object is a regular `ggplot`,
so it can be extended with `+` in the usual way.

```{r ggnet_basic}
bip_ggnet(hr.net, hr)
```

Colours for the two modes are controlled with the `palette` argument (its
default is `c(A = "grey", P = "gold")`). You can override it and layer on extra
`ggplot2` geoms — here we colour by mode and add node labels.

```{r ggnet_styled, message = FALSE, warning = FALSE}
col <- c("P" = "#FC9272", "A" = "#9ECAE1")
bip_ggnet(hr.net, hr,
          size = 7, shape = "mode", color = "mode",
          palette = col, layout.exp = 0.25) +
  geom_text(aes(label = network.vertex.names(hr.net)),
            color = "black", size = 3) +
  theme(legend.position = "none")
```

## The railway layout

`bip_railway()` uses the classic two-parallel-columns layout familiar from the
`bipartite` package, with each mode on its own axis.

```{r railway, message = FALSE, warning = FALSE}
bip_railway(hr, label = TRUE)
```

## References

Bascompte, J. & Jordano, P. (2014) *Mutualistic Networks*. Princeton
University Press, Princeton, NJ.

Pocock, M.J.O. *et al.* (2016) The visualisation of ecological networks, and
their use as a tool for engagement, advocacy and management. *Advances in
Ecological Research* **54**, 41–85.

```{r session_info}
sessionInfo()
```
